---
title: "Spatial lab"
author: "lz"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## {.tabset}

### Packages and Data

Install the following packages and set the directory where data are located:

```{r pkgs, message=F, warning=F}
library(ggplot2)
if (!require(dplyr)) {install.packages("dplyr"); library(dplyr)}
# plotting spatial data
if (!require(leaflet)) {install.packages("leaflet"); library(leaflet)}  
# to save html maps that are interactive
if (!require(htmlwidgets)) {install.packages("htmlwidgets"); library(htmlwidgets)}  
# handle simple features objects
if (!require(sf)) {install.packages("sf"); library(sf)} 

```


- We create a set of geospatial points representing cities and their populations.


```{r data, message=F, warning=F}
cities <- read.csv(textConnection("
city,lat,lon,pop
Boston,42.3601,-71.0589,645966
Hartford,41.7627,-72.6743,125017
New York City,40.7127,-74.0059,8406000
Philadelphia,39.9500,-75.1667,1553000
Pittsburgh,40.4397,-79.9764,305841
Providence,41.8236,-71.4222,177994
"))
cities
```

- External data 

- `pm25_ca_2018.csv`
- `pm25_sites.csv`
- `shapes` <- This is a folder


### Mapping Point Data

Let's plot the cities on a map using the leaflet package.

```{r pointsmap1}
  leaflet(cities) %>% 
  addTiles() %>%
  addCircles(lng = ~lon, lat = ~lat, 
             color='firebrick', fillOpacity=1, opacity=1, radius=700)

```

Try changing the basemap with different "provider tiles" using addProviderTiles option (e.g., Wikimedia, CartoDB.Positron) https://leaflet-extras.github.io/leaflet-providers/preview/

```{r pointsmap2}

leaflet(cities) %>% 
  addProviderTiles('Wikimedia') %>% 
  addCircles(lng = ~lon, lat = ~lat,
             fillOpacity=1, opacity=1, radius=700)

```


Now let's make the circle size proportional to the population of each city

```{r pointsmap3}

leaflet(cities) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(lng = ~lon, lat = ~lat, weight = 1, 
             color = "purple", opacity = 1,
             radius = ~sqrt(pop) * 40, popup = ~city)

```


Finally let's colour the circle size differently by population, add a legend, and add popups with labels.

```{r pointsmap4}

pop.pal = colorNumeric(c('blue','brown','brown'), domain=cities$pop)
leaflet(cities) %>% 
  addTiles() %>%
  addCircles(lng = ~lon, lat = ~lat, weight = 20, 
             color = ~pop.pal(pop), opacity = 1,
             label=~paste0(city, ': ', pop)) %>%
  addLegend(position = "bottomright", pal = pop.pal, values = ~pop,
            title="Population",opacity=1)

```

Hover your cursor over the points to see the label appear.


Now let's import some external geospatial data (PM2.5 concentrations from monitoring stations) and map them.

- Data source: [EPA PM2.5](https://www.epa.gov/outdoor-air-quality-data/download-daily-data)


```{r pmdat}
pm25 = read.csv("pm25_ca_2018.csv")
pm25.sites = read.csv("pm25_sites.csv")
pm25 %>% head()
pm25.sites %>% head()
```

We will look at the days where there were major forest fires in November 2018.   saveWidget() will save the interactive plot to an *.html file.

```{r pm-maps}
pm25.pal = colorNumeric(c('darkgreen','goldenrod','brown'),
                        domain=pm25$pm25)
dat <- merge(pm25 %>% filter(date=='11/10/18'), pm25.sites, by='site_id') %>% 
  st_as_sf(coords=c('longitude','latitude'), crs=4326) 
 
m =  leaflet(dat) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(label=~paste0(pm25, ' ug/m3'), color=~pm25.pal(pm25),
             opacity=1, fillOpacity=1, radius=500) %>%
  addLegend('bottomleft', pal=pm25.pal, values=pm25$pm25,
            title='PM2.5 (ug/m3)', opacity=1)
m
saveWidget(m, file=c("ca_map.html"))
```

### Mapping Areal Data

Typically come in Polygon geometries. They aggregate/summarize values (e.g., population, income) over geographic divisions (e.g., countries, states, counties, census blocks, etc.).

Areal data are most often stored as shapefiles (.shp). Shapefiles usually come in a set of at least 4 files:

- `.shp`: main shapefile
- `.shx`: spatial index
- `.dbf`: dBase file which stores the non-spatial columns
- `.prj`: definition of the projection of the spatial data

For example, the shapefile `ca_counties.shp` and its related files consist of polygons for California counties. We read shapefiles with the `st_read()` function from the `sf` package:


```{r shp}
ca.counties = st_read("shapes/ca_counties.shp")
ca.counties
ca.counties %>% class()
```

```{r geom-polygons}

ca.counties %>% 
  leaflet() %>% 
  addProviderTiles('OpenStreetMap') %>% 
  addPolygons(weight=1, fillOpacity=.25, label=~county)
```

An `sf` object is also a `data.frame` so most of the time typically `data.frame` operations work the same way. `geometry` is the unique column to look out for.

You can also create your own `sf` objects, most typically for Point data when given coordinates (longitude and latitude) of the features, using the `st_as_sf()` function.



```{r areal-data}
pop.pal = colorNumeric(c('beige','brown','brown'),
                       domain=ca.counties$population)
ca.counties %>% 
  leaflet() %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addPolygons(fillColor=~pop.pal(population), weight=.5, fillOpacity=.9,
              label=~paste0(county, ': ', population))
```

### Mapping Lines/Polylines

Shapefiles can also represent lines. For example, we can look at the LA Metro Rail lines

```{r geom-lines}
metro.rails = st_read("shapes/metro_lines.shp")
line.pal = colorFactor(c('blue','cornflowerblue','darkgoldenrod','darkgreen','firebrick'),
                       domain=metro.rails$line)
metro.rails %>%
  leaflet() %>% 
  # addProviderTiles('Wikimedia') %>% 
  addProviderTiles(providers$Esri.WorldStreetMap) %>% 
  addPolylines(color=~line.pal(line), label=~line, opacity=.9)
```



Try your own points of interest, different basemaps, different colour schemes, submit some maps.


