---
title: "LA's BEST -- Hypothesis Testing - part I"
author: "Kim Siegmund"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if(!require("genefilter")) {BiocManager::install("genefilter")}
if(!require("ggplot2")) {install.packages("ggplot2")}
library(genefilter)
library(ggplot2)
```

# {.tabset}

## Simulation

We want to test for difference in mean blood pressure levels in smokers and non-smokers. I  will  assume blood pressure level is normally distributed with

* population mean blood pressure in smokers is 122 mm Hg 

* population mean blood pressure in non-smokers is 120 mm Hg

* population standard deviation is 5

We can write this symbolically like this:

Ys~N(mean=122, sd=5)
Yns~N(mean=120, sd=5)

where Ys is the distribution of blood pressure in smokers, and Yns is distribution in non-smokers.

Let's visualize what these distributions look like by picking a random sample of 50 blood pressure measurements from the distribution in smokers and 50 from the distribution in nonsmokers. We can do it in R like this:

```{r sim_data}
#set.seed(35)
n1 = n2 = 50
ys  <- rnorm(n = n1,mean = 122, sd = 5)   
yns <- rnorm(n = n2,mean = 120, sd = 5)
y <- c(ys,yns)
```

The last statement combines the two sets of measures into a single variable y, with the 50 measurements in smokers first, and followed by the 50 measurements in non-smokers.

Now I need to create a smoking variable that identifies the first 50 measurements as being from smokers  and the second 50 from non-smokers. I will use the function rep(x,n=50) to create 50 copies of x, and then create a 'factor' variable for my analysis.  Here's what that looks like:
```{r smoke}
smoke <- c(rep("smokers",n1),rep("nonsmokers",n2))
smoke <- factor(smoke,
                levels=c("smokers","nonsmokers"))
```

Now we can visualize the sample distributions for these two groups.
```{r figures}
dat <- cbind.data.frame(y,smoke)
ggplot(data=dat, aes(x=y, fill=smoke)) +
    geom_density(adjust=1.5, alpha=.4) +
     xlim(100,140)
```

Does the average blood pressure in smokers appears to be higher?  Why or why not?

Sometimes we might use a boxplot to compare distributions.

```{r boxplot}
boxplot(y~smoke)
```

Let's also create a dataset that doesn't have a difference in means. I'm going to do this by splitting the data into two random subsets of 100. I do this by shuffling ('permuting') the smoking values. 

```{r permutex}
rsmoke <- smoke[sample(n1+n2)]
head(cbind.data.frame(smoke,rsmoke))
```

Now, the 50 blood pressure measurements from 'smokers' are drawn from the entire y distribution and no longer reflect the first 50 rows in the data. 

```{r density-permutedx}
dat <- cbind.data.frame(y,rsmoke)
ggplot(data=dat, aes(x=y, fill=rsmoke)) +
    geom_density(adjust=1.5, alpha=.4) +
     xlim(100,140)
```

Does this figure show 2 groups with similar distributions? 


## No difference in means (Ho)

Let's test difference in blood pressure for two groups with similar distributions. We use the permuted (shuffled) smoking variable for this. 

**Q1:** What is your null hypothesis?

**Q2:** What do you expect you will see?  (Will you reject Ho?)

```{r 2gpttest-null}
t.test(y~rsmoke,var.equal=T)
```

Is there a difference in average blood pressure?

CONCLUSION:



## Difference in means (Ha)

Now let's test for a difference in blood pressure for the data original data of 50 smokers and 50 non-smokers.

**Q:** What do you expect you will see?  (Will you reject Ho?)

```{r 2gpttest-alt}
t.test(y~smoke,var.equal=T)
```

Is there a difference in average blood pressure?

CONCLUSION:


Let's return to the main room and see if everyone got the same result... (n=18 independent studies).



## Session Info
```{r sI}
sessionInfo()
```
