---
title: "LA's BeST — Lab 1: Overfitting and the Train/Test Split"
author: ""
date: ""
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Overview

This lab demonstrates two related but distinct ideas from the lecture.

**Issue 1 — Evaluation bias:** evaluating a model on the same data used to fit it gives an overly optimistic picture of how well it will generalize. This is true for *any* model, regardless of its complexity. A held-out test set is the remedy.

**Issue 2 — Overfitting:** a model that is too complex will fit the noise in the training data, not just the underlying signal. This makes the train/test gap larger and more damaging — the model does not just get evaluated optimistically, it actually performs worse on new data.

These are separate problems with separate solutions. Parts 2 and 3 illustrate Issue 1. Part 4 illustrates Issue 2.

**By the end of this lab you will have:**

1. Loaded the BodyFat data and explored its structure
2. Split the data into training and test sets
3. Fit a simple linear regression and seen that training performance overestimates test performance even for a well-specified model (Issue 1)
4. Seen how the train/test gap widens as model complexity increases (Issue 2)
5. Understood why both issues together motivate the need for a separate validation set

You will run pre-written code chunks first to build up the tools you need. Then you will apply those tools yourself.

**Understand what each piece of code does and why — don't move on until it makes sense. Focus on the results and what they mean, not on memorizing the R syntax.**

---

# Part 1: Setup (run as-is)

## 1.1 Load the data

```{r load-data}
setwd('/Users/jpl/Library/CloudStorage/GoogleDrive-lewinger@usc.edu/My Drive/Teaching/Machine Learning/Datasets')

bodyfat <- read.csv("bodyfat.csv")

# Remove one erroneous entry with BodyFat = 0
bodyfat <- bodyfat[bodyfat$BodyFat > 0, ]

# How many observations and variables do we have?
dim(bodyfat)

# What does the data look like?
head(bodyfat)
```

**Variable names:** `BodyFat` is the outcome. The remaining columns are predictor variables — age, weight, height, and a series of body circumference measurements.

---

## 1.2 Helper functions

These two functions compute the performance metrics we discussed in the lecture. Run this chunk now — you will use these functions throughout the lab.

```{r helper-functions}
# R-squared
R2 <- function(observed, predicted) {
  ss_res <- sum((observed - predicted)^2)
  ss_tot <- sum((observed - mean(observed))^2)
  1 - ss_res / ss_tot
}

# Root Mean Square Error
RMSE <- function(observed, predicted) {
  sqrt(mean((observed - predicted)^2))
}
```

---

## 1.3 Create the train/test split

```{r split}
set.seed(301)

n <- nrow(bodyfat)
trainset <- sample(1:n, size = floor(0.7 * n))

bodyfat_train <- bodyfat[trainset, ]
bodyfat_test  <- bodyfat[-trainset, ]

cat("Training set size:", nrow(bodyfat_train), "\n")
cat("Test set size:    ", nrow(bodyfat_test), "\n")
```

---

## 1.4 Explore the training data

```{r explore}
par(mar = c(5, 5, 3, 1))
hist(bodyfat_train$BodyFat,
     col = "steelblue3", border = "white",
     xlab = "Body fat (%)", main = "Distribution of body fat % (training set)",
     cex.lab = 1.3, cex.axis = 1.2, cex.main = 1.3)
```


> **Question 1.1:** What range of body fat % would you consider typical for this group? Are there any values that seem surprisingly high or low?


<br>


```{r explore2}
par(mar = c(5, 6, 3, 1))
plot(bodyfat_train$BodyFat ~ bodyfat_train$Abdomen,
     pch = 16, cex = 1.3, col = "red4",
     xlab = "Abdominal circumference (cm)", ylab = "Body fat (%)",
     main = "Body fat vs. abdomen (training set)",
     cex.lab = 1.3, cex.axis = 1.2, cex.main = 1.3)
```
<br>

> **Question 1.2:** If you drew a line through these points, would it fit perfectly? What does the scatter around the line tell you?


> **Question 1.3:** Two men have the same abdominal circumference but very different body fat percentages. What might explain that?


---

# Part 2: Simple model — abdomen only (run as-is)

**Issue 1 — Evaluation bias.** The model we fit here — body fat predicted from abdominal circumference alone — is a reasonable, well-specified model. It is not too complex. Yet you will see that its training performance still overstates how well it predicts new subjects. The train/test gap you observe here has nothing to do with overfitting; it is simply because least squares always minimizes error on the training data, making training performance an optimistic estimate by construction.

## 2.1 Fit the model on training data

```{r fit-simple}
lm_simple <- lm(BodyFat ~ Abdomen, data = bodyfat_train)
summary(lm_simple)
```

> **Note:** In a statistics course you would carefully examine the coefficients, p-values, and confidence intervals in the output above. In machine learning, we typically don't — our goal is prediction, not interpretation. The only number from `summary()` we care about here is the training $R^2$ (shown as "Multiple R-squared"). We will compute it more explicitly in the next chunk.


## 2.2 Training performance

```{r train-perf}
pred_train_simple <- fitted(lm_simple)

R2_train_simple   <- R2(bodyfat_train$BodyFat, pred_train_simple)
RMSE_train_simple <- RMSE(bodyfat_train$BodyFat, pred_train_simple)

cat("--- Simple model: TRAINING performance ---\n")
cat("R-squared:", round(R2_train_simple, 3), "\n")
cat("RMSE:     ", round(RMSE_train_simple, 2), "% body fat\n")
```


```{r plot-train-fit}
par(mar = c(5, 6, 3, 1))
plot(bodyfat_train$BodyFat ~ bodyfat_train$Abdomen,
     pch = 16, cex = 1.3, col = "red4",
     xlab = "Abdominal circumference (cm)", ylab = "Body fat (%)",
     main = "Simple model fitted on training data",
     cex.lab = 1.3, cex.axis = 1.2, cex.main = 1.2,
     xlim = c(65, 135), ylim = c(0, 50))
abline(lm_simple, col = "steelblue3", lwd = 3)
```

## 2.3 Test performance

```{r test-perf}
pred_test_simple <- predict(lm_simple, newdata = bodyfat_test)

R2_test_simple   <- R2(bodyfat_test$BodyFat, pred_test_simple)
RMSE_test_simple <- RMSE(bodyfat_test$BodyFat, pred_test_simple)

cat("--- Simple model: TEST performance ---\n")
cat("R-squared:", round(R2_test_simple, 3), "\n")
cat("RMSE:     ", round(RMSE_test_simple, 2), "% body fat\n")
```

```{r compare-simple}
cat("=== Simple model: Train vs. Test ===\n")
cat(sprintf("%-10s  R2 = %.3f   RMSE = %.2f%%\n", "Train:", R2_train_simple, RMSE_train_simple))
cat(sprintf("%-10s  R2 = %.3f   RMSE = %.2f%%\n", "Test:",  R2_test_simple,  RMSE_test_simple))
```


> **Question 2.1:** Is the test $R^2$ higher or lower than the training $R^2$? Is the test RMSE higher or lower than the training RMSE? What does the difference tell you?



```{r plot-test-fit}
par(mar = c(5, 6, 3, 1))
plot(bodyfat_test$BodyFat ~ bodyfat_test$Abdomen,
     pch = 16, cex = 1.3, col = "red3",
     xlab = "Abdominal circumference (cm)", ylab = "Body fat (%)",
     main = "Simple model evaluated on test data",
     cex.lab = 1.3, cex.axis = 1.2, cex.main = 1.2,
     xlim = c(65, 135), ylim = c(0, 50))
abline(lm_simple, col = "steelblue3", lwd = 3)
```

```{r plot-pred-obs}
par(mar = c(5, 6, 3, 1))
plot(pred_test_simple ~ bodyfat_test$BodyFat,
     pch = 16, cex = 1.3, col = "red3", asp = 1,
     xlab = "Observed body fat (%)", ylab = "Predicted body fat (%)",
     main = "Predicted vs. observed (test set)",
     cex.lab = 1.3, cex.axis = 1.2, cex.main = 1.2)
abline(0, 1, col = "steelblue3", lwd = 3)
```

> **Question 2.2:** In the predicted vs. observed plot, what would a perfect model look like? How close are we?


---

# Part 3: A richer model — your turn

## 3.1 Fit a model with multiple predictors

```{r fit-multi}
lm_multi <- lm(BodyFat ~ Abdomen + Age + Weight + Neck + Chest,
               data = bodyfat_train)
summary(lm_multi)
```

## 3.2 Evaluate the multi-predictor model

```{r eval-multi}
# Step 1: Get predictions on the training set
pred_train_multi <- fitted(lm_multi)

# Step 2: Get predictions on the test set
pred_test_multi <- predict(lm_multi, newdata = _________)   # fill in

# Step 3: Compute R2 and RMSE on the training set
R2_train_multi   <- R2(_________, _________)   # fill in   
RMSE_train_multi <- RMSE(_________, _________)   # fill in 

# Step 4: Compute R2 and RMSE on the test set
R2_test_multi   <- R2(_________, _________)   # fill in      
RMSE_test_multi <- RMSE(_________, _________)   # fill in    

cat("=== Multi-predictor model: Train vs. Test ===\n")
cat(sprintf("%-10s  R2 = %.3f   RMSE = %.2f%%\n", "Train:", R2_train_multi, RMSE_train_multi))
cat(sprintf("%-10s  R2 = %.3f   RMSE = %.2f%%\n", "Test:",  R2_test_multi,  RMSE_test_multi))
```

## 3.3 Compare both models

```{r compare-models}
cat("=== Model comparison: Test set performance ===\n")
cat(sprintf("%-30s  R2 = %.3f   RMSE = %.2f%%\n", "Simple (abdomen only):",
            R2_test_simple, RMSE_test_simple))  
cat(sprintf("%-30s  R2 = %.3f   RMSE = %.2f%%\n", "Multi (5 predictors):",
            R2_test_multi, RMSE_test_multi))     
```


> **Question 3.1:** Which model performs better on the *test* set? Which performs better on the *training* set? Are they the same?


> **Question 3.2:** The multi-predictor model has a higher training $R^2$ than the simple model. Does that automatically mean it will predict better on new subjects? What does the test set tell us?


---

# Part 4: Overfitting — complexity vs. performance

**Issue 2 — Overfitting.** In Parts 2 and 3 you saw a train/test gap even for reasonable models — that is evaluation bias. Now we look at what happens when we keep adding predictors beyond what is useful. The train/test gap does not just persist — it *widens*, because the model starts fitting noise specific to the training data and performs worse on new subjects as a result. This is overfitting: a distinct problem on top of evaluation bias, with complexity as the culprit.

## 4.1 Building up complexity (run as-is)

We build model complexity in two phases:

- **Phase 1 (steps 1–13):** add the 13 real predictors one at a time, starting with abdomen
- **Phase 2 (steps 14 onward):** add squared terms (abdomen², age², ...), then the first two cubic terms

The squared and cubic terms let the model fit curved, non-linear relationships — but also give it enough flexibility to start fitting noise.

> **Note:** We are showing test $R^2$ here purely to see what overfitting looks like — to understand it visually. In a real analysis you would never use the test set to compare or select among models. Question 4.2 asks you to think about exactly that.

```{r complexity-curve}
cont_vars <- c("Abdomen", "Age", "Weight", "Height", "Neck", "Chest",
               "Hip", "Thigh", "Knee", "Ankle", "Biceps", "Forearm", "Wrist")

# Phase 1: linear terms
phase1 <- cont_vars

# Phase 2: squared terms, then first 2 cubic terms
phase2 <- c(paste0("I(", cont_vars, "^2)"),
            paste0("I(", cont_vars[1:2], "^3)"))   # Abdomen^3, Age^3

all_terms <- c(phase1, phase2)
n_linear  <- length(phase1)          # 13 — boundary between phases

R2_train_curve <- numeric(length(all_terms))
R2_test_curve  <- numeric(length(all_terms))

for (k in seq_along(all_terms)) {
  formula_k <- as.formula(
    paste("BodyFat ~", paste(all_terms[1:k], collapse = " + "))
  )
  fit_k <- lm(formula_k, data = bodyfat_train)
  R2_train_curve[k] <- R2(bodyfat_train$BodyFat, fitted(fit_k))
  R2_test_curve[k]  <- R2(bodyfat_test$BodyFat,  predict(fit_k, newdata = bodyfat_test))
}
```

```{r plot-complexity}
n_steps <- length(all_terms)

par(mar = c(5, 6, 4, 1))
plot(1:n_steps, R2_train_curve,
     type = "l", lwd = 2.5, col = "steelblue3",
     xlab = "Model complexity (terms added)", ylab = expression(R^2),
     main = "Train vs. Test R² as model complexity increases",
     cex.lab = 1.3, cex.axis = 1.2, cex.main = 1.2,
     ylim = c(min(R2_test_curve) - 0.05, 1.0), xaxt = "n")
axis(1, at = 1:n_steps, labels = 1:n_steps, cex.axis = 0.9)
lines(1:n_steps, R2_test_curve, lwd = 2.5, col = "red4")

# Phase boundary
abline(v = n_linear + 0.5, col = "gray50", lty = 2, lwd = 1.5)

# R² = 0 reference line
abline(h = 0, col = "gray70", lty = 3, lwd = 1.5)

# Phase labels
mtext("Linear terms", side = 3, at = n_linear / 2,        cex = 0.9, col = "gray40")
mtext("Polynomial terms", side = 3, at = n_linear + (n_steps - n_linear) / 2,
      cex = 0.9, col = "gray40")

legend("topright",
       legend = c("Training R²", "Test R²"),
       col = c("steelblue3", "red4"), lwd = 2.5, cex = 1.1)
```

> **Question 4.1:** Does training $R^2$ ever decrease as you add more predictors or polynomial terms?

> **Question 4.2:** Describe what happens to test $R^2$ across the two phases. What does a negative test $R^2$ mean in practice?

> **Question 4.3:** We used the test set to draw this plot. Why does that mean we could not validly use this plot to select the best model and then report its test $R^2$ as our performance estimate?


---

# Part 5: Your own model — your turn

## 5.1 Fit your own model

```{r show-vars}
names(bodyfat)
```

```{r your-model}
# Replace the ... with your chosen predictors
lm_yours <- lm(BodyFat ~ ..., data = bodyfat_train)

summary(lm_yours)
```


## 5.2 Evaluate your model on the test set

```{r eval-your-model}
pred_train_yours <- fitted(lm_yours)
pred_test_yours  <- predict(lm_yours, newdata = bodyfat_test)

R2_train_yours   <- R2(bodyfat_train$BodyFat, pred_train_yours)
RMSE_train_yours <- RMSE(bodyfat_train$BodyFat, pred_train_yours)

R2_test_yours   <- R2(bodyfat_test$BodyFat, pred_test_yours)
RMSE_test_yours <- RMSE(bodyfat_test$BodyFat, pred_test_yours)

cat("=== Your model: Train vs. Test ===\n")
cat(sprintf("%-10s  R2 = %.3f   RMSE = %.2f%%\n", "Train:", R2_train_yours, RMSE_train_yours))
cat(sprintf("%-10s  R2 = %.3f   RMSE = %.2f%%\n", "Test:",  R2_test_yours,  RMSE_test_yours))
```

## 5.3 Final comparison

```{r final-comparison}
cat("=== Final comparison: Test set R² ===\n")
cat(sprintf("  Simple model (abdomen only):  R2 = %.3f\n", R2_test_simple))
cat(sprintf("  Multi model (5 predictors):   R2 = %.3f\n", R2_test_multi))
cat(sprintf("  Your model:                   R2 = %.3f\n", R2_test_yours))
```

> **Question 5.1:** How does your model compare on the test set? Did adding your chosen predictors help?


> **Question 5.2:** Suppose you tried 5 different combinations of predictors, picked the one with the best test $R^2$, and reported that as your model's performance. What would be the right way to proceed instead?


---

# Summary

In this lab you have seen two distinct ideas:

**Issue 1 — Evaluation bias (Parts 2–3):**

- Training performance always overestimates how well a model will generalize — for *any* model, not just complex ones
- A held-out test set is the only way to get an honest estimate of generalization performance
- The test set must be treated as **sacred**: looked at once, after all modeling decisions are made

**Issue 2 — Overfitting (Part 4):**

- Adding complexity always improves training $R^2$, but test $R^2$ eventually peaks and declines
- Beyond a certain point, a more complex model does not just get evaluated optimistically — it actually predicts new subjects *worse*
- Controlling complexity is a separate problem from honest evaluation, and requires a separate solution: a validation set (or cross-validation)

Both issues together are why we need the train / validation / test framework introduced at the end of the lecture — and why cross-validation, decision trees, and random forests are the natural next steps in Lecture 2.
