---
title: "R Markdown Basics"
author: "Aria Fu"
date: "2026-06-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Markdown
A widely used formatting language: Github, Jupyter notebook.

A simple way to format text using plain characters — `##` makes a heading, `**bold**` makes **bold**, backticks make code. No clicking buttons like in Word.



## Types of Chunks

An (optional) header surrounded by `---`

text mixed with simple text formatting

R code chunks surrounded by ` ``` `

### Code Chunk

Keyboard shortcuts: `Cmd+Option+I` (Mac) or `Ctrl+Alt+I` (Windows)

#### Code Chunk Options

* `include = FALSE` prevents code and results from appearing in the finished file. R Markdown still runs the code in the chunk, and the results can be used by other chunks.

* `echo = FALSE` prevents code, but not the results from appearing in the finished file. This is a useful way to embed figures.

* `message = FALSE` prevents messages that are generated by code from appearing in the finished file.

* `warning = FALSE` prevents warnings that are generated by code from appearing in the finished.

* `eval = FALSE` Show code but don't run it


## Review of Factor Data Structure

```{r factor review, echo = TRUE, warning=FALSE}
race <- c("White", "Asian", "Hispanic", "Black")
race_obs <- c("White", "Asian", "Hispanic", "Black", "Hispanic", "Black", "Asian", "Asian")
race_obs_f <- factor(race_obs)

print(race_obs)
print(levels(race_obs_f))
print(race_obs_f)
```
## Functions

A function is just a reusable piece of code that takes something in, does something to it, and gives something back.
```{r}
mean(x = c(1, 2, 3, 4, 5))
```

* `mean` — the function name
* `()` — the parentheses are where you pass things in
* `x = c(1,2,3,4,5)` — the argument, the input you're giving it

#### Build your own function
```{r}
my_function <- function(argument) {
  # do something
  return(result)
}
```

```{r}
rectangle_area <- function(width, height) {
  area <- width * height
  return(area)
}

rectangle_area(5, 10)
rectangle_area(height = 10, width = 5)

rec_area <- rectangle_area(6, 12)
rec_area
```

## If-else

#### Basic Structure
```{r, eval=FALSE}
if (condition) {
  # do this if TRUE
} else {
  # do this if FALSE
}
```

Simple example
```{r}
x <- 10

if (x > 5) {
  print("x is greater than 5")
} else {
  print("x is 5 or less")
}
```

## R Packages


A package is a collection of functions someone else wrote and shared, so you don't have to build everything from scratch. There are over 20,000 packages on CRAN (R's official package repository) covering everything from data cleaning to machine learning to making maps.


#### Install Packages
```{r message=FALSE}
# install.packages("dplyr")   # downloads it once
library(dplyr)   # loads it for this session
```

## Use dplyr to do data manipulation

`dplyr` pipeline:

`dplyr` makes data manipulation in R read almost like plain English. Instead of nesting square brackets like in base R, we chain together simple verbs using the pipe `|>` or ` %>% ` The five core verbs are `filter()` (keep rows), `select()` (keep columns), `mutate()` (add or change a column), `arrange()` (sort rows), and `summarise()` (collapse rows into summary statistics, usually paired with `group_by()`). Each step in the pipeline takes the output of the previous step as its input, so a full pipeline reads top to bottom like a sentence describing exactly what's being done to the data.

Note: You may have seen %>% in older tutorials (from the magrittr package).
The native pipe `|>` was added in R 4.1 and behaves identically for most use cases.
Either is fine; `|>` is the modern default.


#### Read Data
```{r}
mtcars1 <- readRDS("/Users/yubofu/Desktop/USC Course/LAbest_2026/mtcars1.rds")

# mtcars2 <- readRDS("mtcars2.rds")

mtcars3 <- read.csv("mtcars.csv") # read CSV file
```




##### `filter()` function for selecting rows

```{r}
cyl4 <- filter(mtcars, cyl == 4) # filter rows with cyl=4
head(cyl4)
```

```{r}
# Multiple conditions — use comma (AND) or | (OR)
cyl4_mpg25 <- filter(mtcars, cyl == 4, mpg > 25)
cyl4_or_mpg25 <- filter(mtcars, cyl == 4 | mpg > 25)
```

##### `select()`: select columns
```{r}
# pick only the columns we care about
subdata1 <- select(mtcars, mpg, cyl, hp, wt)

# drop a column with minus sign
subdata2 <- select(mtcars, -qsec, -vs)
```


##### `mutate()`: add / change columns

```{r}
data3 <- mutate(mtcars, kpl = mpg * 0.425) # create a new dataset with additional kpl col
head(data3)
```
```{r}
mtcars %>%
  mutate(kpl = mpg * 0.425,
         cyl_f = factor(cyl)
         )  %>%
  select(mpg, cyl_f, kpl) %>%
  head()
```


##### Rename columns
```{r}
# rename multiple columns: rename(new_name = old_name)
mtcars %>%
  rename(
  miles_per_gallon = mpg,
  horsepower  = hp
    )
```

##### Order Rows
```{r}
## arrange() — sort rows
# Sort by mpg, ascending (default)
mtcars |>
  arrange(mpg) |>
  head()

# Descending: wrap with desc()
mtcars |>
  arrange(desc(mpg)) |>
  head()
```


##### summarise() + group_by() 
```{r}
# Average mpg across all cars
mtcars  %>%
  summarise(avg_mpg = mean(mpg))

# Average mpg BY cylinder group — this is the power combo
mtcars %>%
  group_by(cyl) %>%
  summarise( avg_mpg = mean(mpg),
             count = n()
            )
```

Full pipeline: which automatic cars have the best efficiency?
```{r}
best_auto <- mtcars |>
  filter(am == 1) |> # automatic only
  select(mpg, cyl, hp, wt) |> # keep relevant columns
  mutate(kpl = mpg * 0.425) |> # add kpl
  arrange(desc(mpg))

head(best_auto)
```
## Plotting

#### Basic plotting

Quick & useful for a fast look at you data
```{r}
# Histogram: distribution of mpg
hist(mtcars$mpg,
     main = "Distribution of fuel efficiency",
     xlab = "Miles per gallon",
     col = "steelblue",
     breaks = 10)

# Scatter plot: weight vs mpg
plot(mtcars$wt, mtcars$mpg,
     main = "Weight vs MPG",
     xlab = "Weight (1000 lbs)",
     ylab = "Miles per gallon",
     pch = 19,
     col = "steelblue")

# Boxplot: mpg by cylinder count
boxplot(mpg ~ cyl, data = mtcars,
        main = "MPG by cylinders",
        xlab = "Cylinders",
        ylab = "MPG",
        col = c("#9FE1CB", "#5DCAA5", "#1D9E75")
        )
```

#### ggplot: making publication quality graphics. (reports, journal papers)
```{r}
# install.packages("ggplot2")
library(ggplot2)

# ggplot works in layers: data → aesthetics → geometry
ggplot(mtcars,
       aes(x = wt, y = mpg, color = factor(cyl))
       ) +
  geom_point(size = 3) +
  labs( title = "Car weight vs fuel efficiency",
        x = "Weight (1000 lbs)",
        y = "Miles per gallon",
        color = "Cylinders"
        ) +
  theme_minimal()
```

