---
title: "spatial1"
author: "Meredith Franklin"
date: "7/1/2021"
output: html_document
---

# {.tabset}  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup {.tabset}

### Packages and Data

Install the following packages and set the directory where data are located:

```{r pkgs, message=F, warning=F}
library(dplyr)
library(ggplot2)
library(leaflet) # plotting spatial data
library(htmlwidgets) # to save html maps that are interactive
library(sf) # handle simple features objects
library(ggmap) # for querying google API for geocoding

```

- We create a set of geospatial points representing cities and their populations.


```{r data, message=F, warning=F}
cities <- read.csv(textConnection("
city,lat,lon,pop
Boston,42.3601,-71.0589,645966
Hartford,41.7627,-72.6743,125017
New York City,40.7127,-74.0059,8406000
Philadelphia,39.9500,-75.1667,1553000
Pittsburgh,40.4397,-79.9764,305841
Providence,41.8236,-71.4222,177994
"))
```

### Mapping

Let's plot the cities on a map using the leaflet package.

```{r pointsmap1}
  leaflet(cities) %>% 
  addTiles() %>%
  addCircles(lng = ~lon, lat = ~lat, 
             color='firebrick', fillOpacity=1, opacity=1, radius=700)

```

Try changing the basemap with different "provider tiles" using addProviderTiles option (e.g. Wikimedia, CartoDB.Positron) https://leaflet-extras.github.io/leaflet-providers/preview/

```{r pointsmap2}

leaflet(cities) %>% 
  addProviderTiles('Wikimedia') %>% 
  addCircles(lng = ~lon, lat = ~lat,
             fillOpacity=1, opacity=1, radius=700)

```


Now let's make the circle size proportional to the population of each city

```{r pointsmap3}

leaflet(cities) %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(lng = ~lon, lat = ~lat, weight = 1, 
             color = "purple", opacity = 1,
             radius = ~sqrt(pop) * 40, popup = ~city)

```


Finally let's colour the circle size differently by population, add a legend, and add popups with labels.

```{r pointsmap4}

pop.pal = colorNumeric(c('blue','brown','brown'), domain=cities$pop)
leaflet(cities) %>% 
  addTiles() %>%
  addCircles(lng = ~lon, lat = ~lat, weight = 20, 
             color = ~pop.pal(pop), opacity = 1,
             label=~paste0(city, ': ', pop)) %>%
  addLegend(position = "bottomright", pal = pop.pal, values = ~pop,
            title="Population",opacity=1)

```

Now let's import some external geospatial data (PM2.5 concentrations from monitoring stations) and map them.

- Data source: [EPA PM2.5](https://www.epa.gov/outdoor-air-quality-data/download-daily-data)


```{r pmdat}
pm25 = read.csv(paste0(data.dir, "/pm25_ca_2018.csv"))
pm25.sites = read.csv(paste0(data.dir, "/pm25_sites.csv"))
pm25 %>% head()
pm25.sites %>% head()
```

We will look at the daysfires from last fall (November 2018)

```{r pm-maps}
pm25.pal = colorNumeric(c('darkgreen','goldenrod','brown','brown','brown'),
                        domain=pm25$pm25)
merge(pm25 %>% filter(date=='11/10/18'), pm25.sites, by='site_id') %>% 
  st_as_sf(coords=c('longitude','latitude'), crs=4326) %>% 
  leaflet() %>% 
  addProviderTiles('CartoDB.Positron') %>% 
  addCircles(label=~paste0(pm25, ' ug/m3'), color=~pm25.pal(pm25),
             opacity=1, fillOpacity=1, radius=500) %>%
  addLegend('bottomleft', pal=pm25.pal, values=pm25$pm25,
            title='PM2.5 (ug/m3)', opacity=1)
```

### Geocoding

Suppose we have names of locations or street addresses - we can query the Google API to geocode them.

- Google API requires a Developer API key, please use yours if you have one or use mine sparingly! Geocoding a large number of locations comes with a fee.

```{r gapi}
register_google(key = "AIzaSyCME84C5t6Al2SBbYLhNPyuxg3reYwghCc")
```

Let's pick some points of interest and geocode them

```{r geoc}
la_locs <- tibble(location = c("Staples Center, Los Angeles", "University of Southern California" ,"Santa Monica Pier"))
# Geocode locations
geo_la_locs <- geocode(la_locs$location)
# Make a data frame with the locations
la_locs_df <- cbind(la_locs, geo_la_locs)
```

Now map the points, and save the map

```{r geoc-map}
m=leaflet(la_locs_df)%>% 
  addProviderTiles("Esri.WorldImagery") %>%
  addCircles(lng = ~lon, lat = ~lat, weight = 15, opacity=0.9)
m

#saveWidget(m, file=paste0(data.dir, "la_map.html"))

```

Try your own points of interest, different basemaps, different colour schemes, submit some maps.


