---
title: "Multiple Linear Regression"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```
# {.tabset}

## Linear regression with a categorical predictor
Using the cholesterol dataset, creating a categorical variable and using it in a linear regression.

First load the cholesterol data again and plot cholesterol vs age
```{r csv, warning=FALSE}
data.dir<-gsub('/labs', '/data', getwd())
print(data.dir)
################################################################
## change data.dir to the dir where you save the chol.csv file. 
################################################################
chol<-read.csv(paste0(data.dir, '/chol.csv'))
#chol<-read.csv('../3Wed/chol.csv')
names(chol)
dim(chol)
ggplot(chol, aes(x=age,y=chol))+
    geom_point()

```

Let's first create a categorical variable of age by binning age by quartiles (i.e., 4 groups with nearly identical sample sizes: <25%, 25-50%, 50-75%, >75% of the distribution). Later we'll
regress chol on the categorical version of the age variable.

```{r cat}
quantile(chol$age)
chol$agegrps <- cut(chol$age, c(0, 11, 17, 35, 70), labels=c("2-11","12-17","18-35","36-68"))
chol %>% group_by(agegrps) %>% summarize(min(age), max(age))
chol$agegrps <- factor(chol$agegrps)
```

Run the regression with the categorical predictor, `agegrps`. How would you 
interpret the coefficient for `agegrps12-17`? 
```{r mod1}
mod1 <- lm(chol ~ agegrps, data = chol)
summary(mod1)
```

Now we change the reference category to be 36-68 years old. By comparing the coef of `agegrps_new2-11` to the coef of `agegrps36-68` from the last 
model, do you find something interesting?
```{r mod2}
chol$agegrps_new <- relevel(chol$agegrps, "36-68")
chol$agegrps_new <- as.factor(chol$agegrps_new)
mod2 <- lm(chol ~ agegrps_new, data = chol)
summary(mod2)
```
How do you interpret the coefs for `agegrps_new2-11`, `agegrps_new12-17`, and `agegrps_new18-35`? 


## Linear regression with several predictors
Now, we fit the model with more predictors besides age groups. 
```{r mod3}
mod3 <- lm(chol ~ tg + female + agegrps, data = chol)
summary(mod3)

coef<-mod3$coefficients
resids<-mod3$residuals
pred<-mod3$fitted.values
```
How do you interpret the coefs for `agegrps_new2-11`, `agegrps_new12-17`, and `agegrps_new18-35` after we include tg and female? 


We can visualize the residuals using histogram, qqplot, resids vs predicted values by using ggplot2. What assumption about residuals are we checking here by using the histogram? 
```{r hist}
ggplot(data.frame(resids), aes(x = resids)) +
  geom_histogram(binwidth = 20)
```


What assumption about residuals are we checking here by using the histograms with a normal curve?
```{r normal}
ggplot(data.frame(resids), aes(x = resids)) +
  geom_histogram(aes(y =..density..), binwidth = 3) +
  stat_function(fun = dnorm, args = list(mean = mean(resids), 
                                         sd = sd(resids)), col = "red")

```

How about this one?
```{r qqplot}
ggplot(data.frame(resids), aes(sample = resids)) +
  stat_qq() +
  stat_qq_line(col = "red")
```


How about the last one? 
```{r pred}
ggplot(data.frame(resids, pred), aes(x = pred, y = resids)) +
  geom_point() +
  geom_abline(slope = 0, col = "red")
```


We want to see if including X variables are statistically better than 
- Not including any X variables 
- Including only a subset of X variables

```{r Ftest}
anova(lm(chol ~ 1, data=chol), lm(chol ~ tg + age, data=chol))
anova(lm(chol ~ tg, data=chol), lm(chol ~ tg + female, data=chol))
anova(lm(chol ~ tg, data=chol), lm(chol ~ tg + agegrps, data=chol))

anova(lm(chol ~ tg + female, data=chol), 
      lm(chol ~ tg + female + agegrps, data=chol))

anova(lm(chol ~ tg + female + agegrps, data=chol), 
      lm(chol ~ tg + female + agegrps + tg*agegrps, data=chol))

```
What can you conclude from these F tests? Should we include the additional variables? Do we need to include the interaction term?

## Exercises

Now try use `log(chol)` as the outcome variable instead of `chol`. Use the 
following code to generate the new outcome variable. 
```{r ex}
chol$lgchol <- log(chol$chol)
```