---
title:  "<span style=color:#003399> Project tips </span>"
author: "Juan Pablo Lewinger"
date: "6/28/2021"
output: ioslides_presentation
widescreen: true
smaller: true
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## <span style="color:#660066"> Scatterplot matrix </span>{.smaller}

```{r eval=FALSE, fig.height=3.5, fig.width=5, message=FALSE, warning=FALSE, include=TRUE}
library(ggplot2)
library(GGally)
setwd("~/LA's best")
chs = read.csv('CHS_cohortE_final_subset.csv')
chs$sex = factor(chs$male, levels=c(0,1), labels=c('F', 'M'))
ggpairs(chs[, c('fev1', 'sex', 'sulfate', 'nitrate')])

```

```{r echo=FALSE, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
library(ggplot2)
library(GGally)
setwd("~/LA's best")
chs = read.csv('CHS_cohortE_final_subset.csv')
chs$sex = factor(chs$male, levels=c(0,1), labels=c('F', 'M'))
gg = ggpairs(chs[, c('fev1', 'sex', 'sulfate', 'nitrate'), ], progress = FALSE)
print(gg)

```

## <span style="color:#660066"> Project  </span> {.smaller}
- Compute descriptive statistics first, particularly graphs 
- ggpairs does a great job with both univariate and bivariate plots
- Regression models
 + Adjust for demographic and personal characteristics: age, male, race, height, and bmi. Include these variables regardless of significance.
 + Run simple models first (i.e. outcome against each risk factor adjusting for variables above)
 + OK to include race  even if low counts for some categories but probably not ok to stratify by race
 + Adjust by parental education (educ), Home age (HomeBuilt), gas stove in residence, (BaseGasstove),  pet in residence (BasePets), and second hand smoke (ETS_base) if significant - Explore the association between outcomes and pollution variables (one at a time and together)
- Use anova() to assess significance of categorical variables with more than 2 levels


## <span style="color:#660066"> Anova example </span> {.smaller} 
```{R}
fev1_model <- lm(fev1 ~  bmi + race, data=chs)
summary(fev1_model) 
```

## <span style="color:#660066"> Anova example </span> {.smaller} 

```{r}
anova(fev1_model)
```


